Viscous Transport in Eroding Porous Media

### 2019-10-01 docs/transport/erosion_transport.tex
leaving off at start of sec 3.2

%Flow in porous media is further complicated when there is a two-way coupling between the flow and the geometry. For example, the grains of a porous media evolve in time in applications involving
\nick{Flow in porous media is further complicated when boundaries evolve dynamically in response to the fluid flow. This type of coupling between geometry and flow occurs, for example, in applications involving} melting~\cite{bec-vis1998, rycroft2016asymmetric, jambon2018singular, favier2019rayleigh, morrow2019moving}, dissolution~\cite{kan-zha-che-he2002, mac2015shape, moo2017, wykes2018self}, deposition~\cite{joh-eli1995,
hewett2018modelling}, biofilm growth~\cite{tan-val-wer2015}, and crack
formation~\cite{cho2019crack}. We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical, hydrological, and
industrial applications~\cite{berhanu2012shape, hewett2017evolution, lachaussee2018competitive, lopez2018cfd, allen2019sde, amin2019role}.  When a porous geometry erodes, certain qualitative characteristics are unveiled that affect transport through the geometry.  For example, an eroded geometry may
contain channels of high porosity, which, though few in number and
modest in volume fraction, transmit a large portion of the flux. This
arrangement results in velocities that vary over several orders of
magnitude~\cite{all-hea-lab-rei2002}.  \nick{Moreover, channelization creates heterogeneous and anisotropic medium properties, which affect the
transport of tracers} such as contaminants~\cite{cve-che-wen1996,
dag1987, kon-bre1978} and heat~\cite{nil-sto1990, ree-sto1995}.

This study consists of two main undertakings: first, high-fidelity
simulations of eroding porous media, and, second, \nick{characterization of tracer transport through the dynamic geometries that result}. Our modeling efforts build on previous work~\cite{ris-moo-chi-she-zha2012, moore2013self, moo2017}, in particular recent numerical methods developed to simulate erosion in the regime of Stokes flow~\cite{qua-moo2018}. We, however, make a few
key improvements to the numerical methods from~\cite{qua-moo2018} to
enable simulations of more realistic, dense suspensions of eroding
grains (Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  To characterize transport through these \nick{configurations}, we examine
coarse-grained variables through statistical analysis of tracer
trajectories.

@@

To compute stable simulations of erosion, we use high-order methods in
both space and time. The time integration method is unchanged from our
previous work~\cite{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities caused by changes in
sign of the shear stress, and we use a stable second-order Runge-Kutta
method applied to the {\thL} coordinates~\cite{hou-low-she1994} of the
eroding grains.  In this work, we \nick{\st{do}} introduce a new quadrature method to resolve dense suspensions of eroding grains.  The accuracy of the
trapezoid rule, which we used in our previous work, is \st{sufficient} \nick{adequate} for bodies that are sufficiently separated~\cite{tre-wei2014}. 
\todo[inline]{I changed the above to `adequate' so that `sufficient' would not be used twice in a row.} 
However, it \nick{loses accuracy whenever \st{two} grains come in close contact}.  One of the earliest works \nick{to address} quadrature methods for nearly-singular integrands was developed by Baker and Shelley~\cite{bak-she1986}, and in recent years, many other near-singular integration schemes have been \nick{followed}~\cite{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016, bea-lai2001, klo-bar-gre-one2013}.  We use a Barycentric quadrature
\nick{method}~\cite{bar2014, bar-wu-vee2015} since it is a non-intrusive
modification of the trapezoid rule, and the error is guaranteed to be
uniformly bounded.  We extend the original quadrature method \nick{to compute the velocity gradient, which is needed to evaluate shear stress and fluid vorticity}.

@@

To characterize the transport, particle trajectories \nick{must} be
computed. Depending on the application, microscale transport can be
modelled as pure advection~\cite{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019},
advection-diffusion~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume trajectories \nick{to be governed by pure advection} (ie.~no diffusion). Therefore, the particle trajectories are
identical to the streamlines, and they can be found, for example, with a
continuous time random walk~\cite{mon-wei1965, den-cor-sch-ber2004,
leb-den-car2008, ber-cor-den-sch2006}.  
\todo[inline]{The logic from the sentence above to the one below seems awkward to me. The above sentence makes me think that a continuous time random-walk will be used in our paper. Then, suddenly, we say we are going to do something else instead. Maybe we can kill the above sentence and dump the references (if you want to keep them) into the earlier sentence that mentions a random walk? Or maybe just a slight rephrasing would make the transition less awkward.}
Instead,  we compute
trajectories $\ss(t)$ that are initialized at $\ss_0$ by solving
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Since there is no stiff diffusive
term in~\eqref{eqn:tracers}, we use a fourth-order explicit Runge-Kutta
time stepping method.  In addition, the Barycentric quadrature rule is
used to accurately compute trajectories that are close to an eroding
grain.

Once \nick{the streamlines are computed, we characterize
transport by analyzing} three different metrics: the tortuosity, the anomalous dispersion, and the pore size distribution. 
The local tortuosity of a streamline \nick{that connects the inlet to the outlet is defined as the streamline's length normalized by the linear inlet-to-outlet distance}.
In porous media, the local tortuosity can be greater than 1.5~\cite{kop-kat-tim1996, mat-kha-koz2008} or even 2~\cite{dud-koz-mat2011}, \nick{depending on several factors such as the porosity.} 
The tortuosity of a geometry is defined by averaging the local tortuosity over all streamlines initialized at the inlet, and the geometry's tortuosity characterizes average particle motions~\cite{hak-com-den2019}.  To characterize spreading, the fluid dispersion is computed by calculating the variance of the streamline lengths. In porous media, this spreading is often
super-dispersive~\cite{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}. \nick{Since anomalous dispersion results from streamlines spending time in both the high and low velocity}
regimes~\cite{ber-sch2001}, it is crucial to accurately resolve  streamlines near grain boundaries, as achieved in this work.
Finally, \nick{the distribution of pore sizes is required to} quantify velocity distributions~\cite{ali-par-wei-bre2017, dea-qua-bir-jua2018}, channelization~\cite{sie-ili-pri-riv-gua2019}, connectivity~\cite{knu-car2005, wes-blo-gra2001}, and to develop network models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}. \nick{Hence, we construct the pore-size distribution throughout the erosion process.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
Our first contribution is extending the Barycentric quadrature rule of
Barnett et al.~\cite{bar-wu-vee2015} \nick{to accurately compute velocity gradients associated with Stokes flow.}
\nick{This new quadrature rule permits accurate and stable simulations of erosion, and, moreover, enables visualization of the vorticity field throughout.}

Our second contribution is computing and analyzing trajectories in
eroded geometries.  These trajectories are used to characterize the
transport by computing the local tortuosity of trajectories, and the
tortuosity and dispersion rate of eroded geometries. By using the
Barycentric quadrature rule and a fourth-order Runge-Kutta method, we
accurately compute trajectories that spend large amounts of time in the
slow velocity regions near the eroding bodies, a necessity to
characterize the flow~\cite{dea-leb-den-tar-bol-dav2013}.  We compare
our tortuosity and dispersion rates with models available in the
literature.  Finally, we analyze the effect of erosion on the pore size
distribution.

\todo[inline]{The above paragraph seems to repeat a lot of what has been mentioned already, but maybe that is okay.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations}
As in our previous work~\cite{qua-moo2018}, we consider two-dimensional
immobile eroding grains. However, this work \nick{does not limit} the proximity of  grains. Compared to our previous work, we are able to
simulate much denser suspensions with a comparable number of
discretization points and without upsampling.  While we do investigate
transport in this work, we only consider passive particles that are
advected by the flow.  To model other physical processes, such as
contaminant transport, we plan to simulate a finite Peclet number
advection-diffusion equation in eroded geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the Paper}
In Section~\ref{sec:formulation}, we summarize the erosion model \nick{that is described in more detail in} previous work~\cite{qua-moo2018}.  In
Section~\ref{sec:DLP}, we recast all the governing equations as layer
potentials defined in both $\RR^2$ and in $\CC$.
Section~\ref{sec:transport} describes measures for characterizing the
geometry, transport, and dispersion. Section~\ref{sec:method} describes
the numerical methods, with special attention paid to the new quadrature
method \nick{for computing} shear stress and vorticity.
Section~\ref{sec:results} presents numerical examples for a variety of
dense packings of bodies.  Finally, concluding remarks are made in
Section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  We only
\nick{briefly summarize the model, since a more detailed description can be found in previous} work~\cite{qua-moo2018}.  We consider flows inside a confined geometry $\Omega$ that contains $M$ eroding bodies with boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$.  
\nick{The boundary of the fluid domain is $\bd \Omega = \Gamma \cup
\gamma_1 \cup \cdots \cup \gamma_M$, where $\Gamma$ is the outer boundary, taken to be a slightly smoothed version of $[-3,3] \times [-1,1]$}. 
\nick{We only place eroding bodies in $[-1,1] \times [-1,1]$ so as to create a buffer region for the flow profile imposed at the inlet to transition to the more complex flow intervening between the bodies}. On each
eroding body, $\nn$ is the normal vector pointing into the body, and
$\ss$ is the unit tangent vector pointing in the counterclockwise
direction.  The normal vector on $\Gamma$ is the outward unit normal.
Neglecting inertial forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the eroding bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here, $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, $\Vn$ is the normal velocity
of $\gamma$, and
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
is the shear stress on $\gamma$.  The strength of $\UU$ is adjusted at
each time step \nick{to achieve} a constant pressure drop across the
channel.  We simulate erosion by alternating between solving the fluid
equations and advancing the grains according to the erosion rate for a
small time step.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Boundary Integral Equation Formulation}
\label{sec:DLP}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Since we require an accurate solution of the incompressible Stokes
equations in complex two-dimensional geometries, we reformulate the
incompressible Stokes equations as a BIE.  This has the advantage that
only the one-dimensional boundary of the domain must be discretized,
and, with an appropriate quadrature formula and fast summation methods,
the result is a high-fidelity numerical simulation of near-optimal
computational complexity.

\subsection{Double-Layer Potential Formulation in $\RR^2$}
Applying the same approach as our previous work~\cite{qua-moo2018}, we
write the velocity $\uu$ as a double-layer potential 
\begin{align}
  \DDD[\eeta](\xx) = \int_{\bd\Omega} D(\xx,\yy) \eeta(\yy)\, ds_\yy = 
  \frac{1}{\pi}\int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) \, ds_\yy, \quad \xx \in \Omega,
  \label{eqn:velocityDLP}
\end{align}
where $D$ is \nick{\st{an} a} kernel of the integral operator, $\rr =
\xx - \yy$, $\rho = \|\rr\|$, $\nn$ is the unit outward normal at $\yy$,
and $\eeta$ is an unknown density function.  The double-layer potential
can not represent all solutions of the incompressible Stokes equations,
so we complete the integral equation formulation by adding the $M$
Stokeslets and rotlets~\cite{pow-mir1987}
\begin{align}
  S[\llambda_\ell,\cc_\ell](\xx) = \frac{1}{4\pi} \left( 
    -\log \rho \II + \frac{\rr \otimes \rr}{\rho^2}
    \right)\llambda_\ell, \quad
  R[\xi_\ell,\cc_\ell](\xx) = \frac{\rr^\perp}{\rho^2} \xi_\ell,
\end{align}
where $\rr = \xx - \cc_\ell$, $\cc_\ell$ is a point inside the
$\ell^{th}$ eroding body, $\rr^\perp = (r_2,-r_1)$, and $\rho =
\|\rr\|$.  Then, for any sufficiently smooth geometry $\Omega$, the
solution of the incompressible Stokes equation with a Dirichlet boundary
condition $\ff$ is
\begin{align}
  \uu(\xx) = \DDD[\eeta](\xx) + 
    \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx), \quad \xx \in \Omega,
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{subequations}
\label{eqn:BIE}
\begin{alignat}{3}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
    \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx) +
    \NN_0[\eeta](\xx), \quad &&\xx \in \bd\Omega, \\
  \llambda_\ell &= \frac{1}{2\pi} \int_{\gamma_\ell} 
    \eeta(\yy)\, ds_\yy &&\ell = 1,\ldots,M, \\
  \xi_\ell &= \frac{1}{2\pi} \int_{\gamma_\ell}
    (\yy - \cc_\ell)^\perp \cdot \eeta(\yy)\, ds_\yy 
    &&\ell = 1,\ldots,M.
\end{alignat}
\end{subequations}
Here, the null space associated with the flux-free condition of $\ff$ is
addressed with the integral operator $\NN_0$ which has the integral
kernel $N_0(\xx,\yy) = \nn(\xx) \otimes \nn(\yy)$, $\xx,\yy \in \Gamma$.
In this work, $\ff$ is the prescribed velocity, $\UU$, on the outer
wall, $\Gamma$, and is zero on the eroding bodies, $\gamma_\ell$,
$\ell=1,\ldots,M$.

>>>>

Flow in porous media is further complicated when boundaries dynamically
evolve in response to the fluid flow. This type of coupling between the
geometry and the flow occurs, for example, in applications involving
melting~\cite{bec-vis1998, rycroft2016asymmetric, jambon2018singular,
favier2019rayleigh, morrow2019moving},
dissolution~\cite{kan-zha-che-he2002, mac2015shape, moo2017,
wykes2018self}, deposition~\cite{joh-eli1995, hewett2018modelling},
biofilm growth~\cite{tan-val-wer2015}, and crack
formation~\cite{cho2019crack}. We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical, hydrological, and
industrial applications~\cite{berhanu2012shape, hewett2017evolution,
lachaussee2018competitive, lopez2018cfd, allen2019sde, amin2019role}.
When a porous geometry erodes, certain qualitative characteristics are
unveiled that affect transport through the geometry.  For example, an
eroded geometry may contain channels of high porosity, which, though few
in number and modest in volume fraction, transmit a large portion of the
flux. This arrangement results in velocities that vary over several
orders of magnitude~\cite{all-hea-lab-rei2002}.  Moreover,
channelization creates heterogeneous and anisotropic medium properties,
which affect the transport of tracers such as
contaminants~\cite{cve-che-wen1996, dag1987, kon-bre1978} and
heat~\cite{nil-sto1990, ree-sto1995}.

This study consists of two main undertakings: first, high-fidelity
simulations of eroding porous media, and, second, characterization of
tracer transport through the resulting eroded geometries. Our modeling
efforts build on previous work~\cite{ris-moo-chi-she-zha2012,
moore2013self, moo2017}, in particular recent numerical methods
developed to simulate erosion in the regime of Stokes flow
regime~\cite{qua-moo2018}. We, however, make key improvements to the
numerical methods from~\cite{qua-moo2018} to enable simulations of more
realistic, dense suspensions of eroding grains
(Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  To
characterize transport through these configurations, we examine
coarse-grained variables through statistical analysis of tracer
trajectories.

@@

To compute stable simulations of erosion, we use high-order methods in
both space and time. The time integration method is unchanged from our
previous work~\cite{qua-moo2018}.  We apply a mild regularization and a
smoothing term to eliminate numerical instabilities caused by changes in
sign of the shear stress, and we use a stable second-order Runge-Kutta
method applied to the {\thL} coordinates~\cite{hou-low-she1994} of the
eroding grains.  In this work, we introduce a new quadrature method to
resolve dense suspensions of eroding grains.  The accuracy of the
trapezoid rule, which we used in our previous work, is adequate for
bodies that are sufficiently separated~\cite{tre-wei2014}.  However, it
loses accuracy whenever grains are in close contact.  One of the
earliest works to address quadrature methods for nearly-singular
integrands was developed by Baker and Shelley~\cite{bak-she1986}, and in
recent years, many other near-singular integration schemes have
followed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  We use a Barycentric quadrature
method~\cite{bar2014, bar-wu-vee2015} since it is a non-intrusive
modification of the trapezoid rule, and the error is guaranteed to be
uniformly bounded.  We extend the original quadrature method to compute
the velocity gradient, which is needed to evaluate shear stress and
fluid vorticity.

@@

To characterize the transport, particle trajectories must be computed.
Depending on the application, microscale transport can be modelled as
pure advection~\cite{dea-qua-bir-jua2018, leb-ded-dav-bou2007,
cve-che-wen1996, puy-gou-den2019},
advection-diffusion~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume trajectories to be governed by pure advection
(ie.~no diffusion), so the particle trajectories are identical to the
streamlines. We compute trajectories $\ss(t)$ that are initialized at
$\ss_0$ by solving the advection equation
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Since there is no stiff diffusive
term in~\eqref{eqn:tracers}, we use a fourth-order explicit Runge-Kutta
time stepping method.  In addition, the Barycentric quadrature rule is
used to accurately compute trajectories that are close to an eroding
grain. We note that an alternative approach to compute trajectories is
to form a distribution of velocities $\uu$ and use a continuous time
random walk~\cite{mon-wei1965, den-cor-sch-ber2004, leb-den-car2008,
ber-cor-den-sch2006}.

Once the streamlines are computed, we characterize transport by
analyzing three different metrics: the tortuosity, the anomalous
dispersion, and the pore size distribution.  The local tortuosity of a
streamline that connects the inlet to the outlet is defined as the
streamline's length normalized by the linear inlet-to-outlet distance.
In porous media, the local tortuosity can be greater than
1.5~\cite{kop-kat-tim1996, mat-kha-koz2008} or even
2~\cite{dud-koz-mat2011}, depending on several factors such as the
porosity. The tortuosity of a geometry is defined by averaging the local
tortuosity over all streamlines initialized at the inlet, and the
geometry's tortuosity characterizes average particle
motions~\cite{hak-com-den2019}.  To characterize spreading, the fluid
dispersion is computed by calculating the variance of the streamline
lengths. In porous media, this spreading is often
super-dispersive~\cite{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}. Since anomalous dispersion results from
streamlines spending time in both the high and low velocity
regimes~\cite{ber-sch2001}, it is crucial to accurately resolve
streamlines near grain boundaries, as achieved in this work.  Finally,
the distribution of pore sizes is required to quantify velocity
distributions~\cite{ali-par-wei-bre2017, dea-qua-bir-jua2018},
channelization~\cite{sie-ili-pri-riv-gua2019},
connectivity~\cite{knu-car2005, wes-blo-gra2001}, and to develop network
models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}. Hence, we
construct the pore-size distribution throughout the erosion process.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
Our first contribution is extending the Barycentric quadrature rule of
Barnett et al.~\cite{bar-wu-vee2015} to accurately compute velocity
gradients associated with Stokes flow. This new quadrature rule permits
accurate and stable simulations of erosion, and, moreover, enables
visualization of the vorticity field throughout.

Our second contribution is computing and analyzing trajectories in
eroded geometries.  These trajectories are formed by combining the
Barycentric quadrature rule and a fourth-order Runge-Kutta method. Then,
they are used to compute tortuosities and dispersion rates of eroded
geometries. Finally, we analyze the effect of erosion on the pore size
distribution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations}
As in our previous work~\cite{qua-moo2018}, we consider two-dimensional
immobile eroding grains. However, this work does not limit the proximity
of  grains. Compared to our previous work, we are able to simulate much
denser suspensions with a comparable number of discretization points and
without upsampling.  While we do investigate transport in this work, we
only consider passive particles that are advected by the flow.  To model
other physical processes, such as contaminant transport, we plan to
simulate a finite Peclet number advection-diffusion equation in eroded
geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the Paper}
In Section~\ref{sec:formulation}, we summarize the erosion model that is
described in more detail in previous work~\cite{qua-moo2018}.  In
Section~\ref{sec:DLP}, we recast all the governing equations as layer
potentials defined in both $\RR^2$ and in $\CC$.
Section~\ref{sec:transport} describes measures for characterizing the
geometry, transport, and dispersion. Section~\ref{sec:method} describes
the numerical methods, with special attention paid to the new quadrature
method for computing the shear stress and the vorticity.
Section~\ref{sec:results} presents numerical examples for a variety of
dense packings of bodies.  Finally, concluding remarks are made in
Section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  We only
briefly summarize the model, and a more detailed description can be
found in previous work~\cite{qua-moo2018}.  We consider flows inside a
confined geometry $\Omega$ that contains $M$ eroding bodies with
boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$.  The boundary of the
fluid domain is $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup
\gamma_M$, where $\Gamma$ is the outer boundary, taken to be a slightly
smoothed version of $[-3,3] \times [-1,1]$.  We only place eroding
bodies in $[-1,1] \times [-1,1]$ so as to create a buffer region for the
flow profile imposed at the inlet to transition to the more complex flow
intervening between the bodies. On each eroding body, $\nn$ is the
normal vector pointing into the body, and $\ss$ is the unit tangent
vector pointing in the counterclockwise direction.  The normal vector on
$\Gamma$ is the outward unit normal.  Neglecting inertial forces, the
governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the eroding bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here, $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, $\Vn$ is the normal velocity
of $\gamma$, and
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
is the shear stress on $\gamma$. We simulate erosion by alternating
between solving the fluid equations nd advancing the grains according to
the erosion rate for a small time step. The strength of $\UU$ is
adjusted at each time step to achieve a constant pressure drop across
the channel.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Boundary Integral Equation Formulation}
\label{sec:DLP}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Since we require an accurate solution of the incompressible Stokes
equations in complex two-dimensional geometries, we reformulate the
incompressible Stokes equations as a BIE.  This has the advantage that
only the one-dimensional boundary of the domain must be discretized,
and, with an appropriate quadrature formula and fast summation methods,
the result is a high-fidelity numerical simulation with near-optimal
computational complexity.

\subsection{Double-Layer Potential Formulation in $\RR^2$}
Applying the same approach as our previous work~\cite{qua-moo2018}, we
write the velocity $\uu$ as a double-layer potential 
\begin{align}
  \DDD[\eeta](\xx) = \int_{\bd\Omega} D(\xx,\yy) \eeta(\yy)\, ds_\yy = 
  \frac{1}{\pi}\int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) \, ds_\yy, \quad \xx \in \Omega,
  \label{eqn:velocityDLP}
\end{align}
where $D$ is the kernel of the integral operator, $\rr = \xx - \yy$,
$\rho = \|\rr\|$, $\nn$ is the unit outward normal at $\yy$, and $\eeta$
is an unknown density function.  The double-layer potential can not
represent all solutions of the incompressible Stokes equations, so we
complete the integral equation formulation by adding the $M$ Stokeslets
and rotlets~\cite{pow-mir1987}
\begin{align}
  S[\llambda_\ell,\cc_\ell](\xx) = \frac{1}{4\pi} \left( 
    -\log \rho \II + \frac{\rr \otimes \rr}{\rho^2}
    \right)\llambda_\ell, \quad
  R[\xi_\ell,\cc_\ell](\xx) = \frac{\rr^\perp}{\rho^2} \xi_\ell,
  \quad \ell=1,\ldots,M,
\end{align}
where $\rr = \xx - \cc_\ell$, $\cc_\ell$ is a point inside the
$\ell^{th}$ eroding body, $\rr^\perp = (r_2,-r_1)$, and $\rho =
\|\rr\|$.  Then, for any sufficiently smooth geometry $\Omega$, the
solution of the incompressible Stokes equation with a Dirichlet boundary
condition $\ff$ is
\begin{align}
  \uu(\xx) = \DDD[\eeta](\xx) + 
    \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx), \quad \xx \in \Omega,
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{subequations}
\label{eqn:BIE}
\begin{alignat}{3}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
    \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx) +
    \NN_0[\eeta](\xx), \quad &&\xx \in \bd\Omega, \\
  \llambda_\ell &= \frac{1}{2\pi} \int_{\gamma_\ell} 
    \eeta(\yy)\, ds_\yy &&\ell = 1,\ldots,M, \\
  \xi_\ell &= \frac{1}{2\pi} \int_{\gamma_\ell}
    (\yy - \cc_\ell)^\perp \cdot \eeta(\yy)\, ds_\yy 
    &&\ell = 1,\ldots,M.
\end{alignat}
\end{subequations}
Here, the null space associated with the flux-free condition of $\ff$ is
addressed with the integral operator $\NN_0$ which is the integral
operator with kernel $N_0(\xx,\yy) = \nn(\xx) \otimes \nn(\yy)$,
$\xx,\yy \in \Gamma$.  In this work, $\ff$ is the prescribed velocity,
$\UU$, on the outer wall, $\Gamma$, and is zero on the eroding bodies,
$\gamma_\ell$, $\ell=1,\ldots,M$.

Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, the
Stokeslets $\llambda_\ell$, and the rotlets $\xi_\ell$, we are able to
compute the deformation tensor, pressure, and vorticity, defined
respectively as
\begin{align}
  \label{eqn:deformationDLP} 
  \ssigma(\xx) &= \int_{\bd\Omega} \mathbf{D}_{\ssigma}(\xx,\yy) 
      \eeta(\yy) ds_{\yy} 
  = \frac{1}{2\pi}\int_{\bd\Omega} \left(
    2\frac{\rr \cdot \nn}{\rho^2} \frac{\rr \cdot \eeta}{\rho^2} \II + 
    \frac{\rr \cdot \eeta}{\rho^4} (\nn \otimes \rr + \rr \otimes \nn) 
    \right. \nonumber \\
    & \qquad\qquad\qquad\qquad\qquad\qquad \left.
    +\frac{\rr \cdot \nn}{\rho^4} (\eeta \otimes \rr + \rr \otimes \eeta) - 
    8\frac{(\rr \cdot \nn)(\rr \cdot \eeta)}{\rho^6}(\rr \otimes \rr)
    \right) \, ds_\yy, \\
%
  p(\xx) &= \int_{\bd\Omega} D_{p}(\xx,\yy) 
      \eeta(\yy) ds_{\yy} 
  = -\frac{1}{\pi}\int_{\bd\Omega} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) \nn \cdot
      \eeta(\yy)\,ds_\yy, \\
%
  \label{eqn:vorticityDLP} 
  \omega(\xx) &= \int_{\bd\Omega} D_{\omega}(\xx,\yy) 
      \eeta(\yy) ds_{\yy} 
  = -\frac{1}{\pi} \int_{\bd\Omega}
    \frac{(\rr \cdot \nn^\perp) + (\rr \cdot \nn)}{\rho^4} 
      (\rr \cdot \eeta) \, ds_\yy.
\end{align}
Since the deformation tensor of the double-layer potential is
discontinuous across $\bd\Omega$, the jump term 
\begin{align}
  \frac{1}{2} \left(\pderiv{\eeta}{\ss} \cdot \ss \right) \left[
    \begin{array}{cc}
      s_x^2 - s_y^2 & 2s_x s_y \\ 2s_x s_y & s_y^2 - s_x^2
    \end{array}
  \right],
  \label{eqn:deformationJump}
\end{align}
must be added to $\ssigma(\xx)$ for $\xx \in \bd\Omega$.  Finally, the
Stokeslets and rotlets do not exhibit any jumps across $\gamma$, so the
associated components of the deformation tensor, pressure, and vorticity
due to the Stokeslets and rotlets are straightforward to
compute~\cite{poz1992}.  Having computed the deformation tensor on
$\gamma$, the shear stress is computed as defined in
equation~\eqref{eqn:shearStress}.
